---
title: "TÜİK Forecasting Project: Non-Domestic Producer Price Index"
author: "Mehmet Zahid Kemik (138722034)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)

required_packages <- c("tuikr", "dplyr", "knitr", "kableExtra", "readxl", "crul")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "tuikr") {
      if (!requireNamespace("remotes", quietly = TRUE)) {
        install.packages("remotes")
      }
      remotes::install_github("emraher/tuikr")
    } else {
      install.packages(pkg)
    }
  }
}

source("R/data_import.R")
source("R/forecasting_methods.R")
source("R/accuracy_measures.R")
source("R/plots.R")

if (!dir.exists("outputs/tables")) dir.create("outputs/tables", recursive = TRUE)
if (!dir.exists("outputs/figures")) dir.create("outputs/figures", recursive = TRUE)
```

# Project Overview

This project forecasts the next monthly observation of the **Non-Domestic Producer Price Index (2010=100)** published by TÜİK. This index tracks import prices in Türkiye and is closely related to exchange-rate pass-through and inflation dynamics.

The analysis uses the `tuikr` package to access TÜİK data. When the SDMX endpoint returns an authorization error, the notebook automatically falls back to the official TÜİK portal table path documented in the `tuikr` package.

# Data Source and TÜİK Connection

```{r data-access}
project_data <- load_project_series(lang = "en")
raw_data <- project_data$raw_data
series_tbl <- project_data$series
target_info <- project_data$target
access_date <- project_data$access_date
meta <- project_data$meta

knitr::kable(data.frame(
  Field = c(
    "TÜİK data set name",
    "TÜİK theme / category",
    "TÜİK table name",
    "tuikr dataflow ID",
    "Selected variable",
    "Data frequency",
    "Time coverage",
    "Latest available observation",
    "Forecast target period",
    "Date of data access",
    "Data access method"
  ),
  Value = c(
    meta$dataset_name,
    meta$theme,
    meta$table_name,
    meta$dataflow_id,
    meta$selected_variable,
    meta$frequency,
    paste(format_period_label(min(series_tbl$period_date)), "to", format_period_label(max(series_tbl$period_date))),
    target_info$latest_observation,
    target_info$forecast_target_period,
    as.character(access_date),
    project_data$access_method
  )
))
```

The data were accessed through the `tuikr` package. The notebook first attempts `tuikr::statistical_data()` using the registered SDMX dataflow ID. If the SDMX service returns an authorization error, it uses `tuikr::statistical_tables()` to locate the official TÜİK portal table and downloads that table programmatically inside R without saving a separate data file to the repository.

Package source: [https://github.com/emraher/tuikr](https://github.com/emraher/tuikr)

```{r show-raw-head}
knitr::kable(head(raw_data, 10))
```

# Use of TÜİK Data in R

After importing the TÜİK table through `tuikr`, the notebook keeps only the **national monthly non-domestic producer price index level**. Rate-of-change indicators and disaggregated sub-indices are excluded so that the selected variable remains a single numeric time series suitable for forecasting.

The main adjustments performed in R are:

1. Selecting the non-domestic producer price index variable rather than monthly or annual change rates.
2. Using `obsTime` as the period variable and converting it to monthly dates.
3. Sorting observations chronologically and removing duplicate periods.
4. Confirming that `obsValue` is numeric and checking for missing observations.

```{r data-prep-summary}
cat("Number of observations:", nrow(series_tbl), "\n")
cat("First period:", format_period_label(min(series_tbl$period_date)), "\n")
cat("Last period:", format_period_label(max(series_tbl$period_date)), "\n")
cat("Any missing values:", any(is.na(series_tbl$value)), "\n")
```

# Exploratory Time Series Analysis

```{r exploratory-plot, fig.cap="Actual monthly Non-Domestic Producer Price Index."}
plot(
  series_tbl$period_date,
  series_tbl$value,
  type = "l",
  lwd = 2,
  col = "#1f4e79",
  xlab = "Period",
  ylab = "Index (2010=100)",
  main = "Non-Domestic Producer Price Index"
)
grid()
```

**Trend:** The series shows a strong upward trend, especially in recent years, which reflects imported inflation pressure in Türkiye.

**Seasonality:** Monthly oscillations are visible. Because the data frequency is monthly, seasonal methods with a 12-month cycle are applicable.

**Random variation:** Short-term fluctuations remain around the trend, indicating that no method fits perfectly without error.

**Missing values / outliers:** The prepared series does not contain missing monthly periods in the filtered national index series. Extreme increases during high-inflation periods are economically meaningful rather than data errors.

# Forecasting Methods Applied

The following sections apply all required forecasting methods. Accuracy is evaluated on in-sample fitted or one-step-ahead values, while the next-period forecast uses the full sample.

## 7.1 Naïve Forecasting

```{r naive}
method_results <- run_all_forecasting_methods(series_tbl)
naive_result <- method_results$naive
tail(data.frame(actual = naive_result$actual, forecast = naive_result$forecast), 5)
calc_accuracy_measures(naive_result$actual, naive_result$forecast)
```

**Interpretation:** The naïve method assumes that the next value equals the most recent observation. It is simple, but it ignores trend and seasonality.

## 7.2 Moving Average

A **3-month window** is used because producer price inflation reacts quickly to recent shocks, while a longer window would over-smooth short-run dynamics.

```{r moving-average}
ma_result <- method_results$moving_average
calc_accuracy_measures(ma_result$actual, ma_result$forecast)
```

**Interpretation:** The moving average smooths recent observations and reduces noise, but it reacts more slowly than exponential smoothing when the trend accelerates.

## 7.3 Weighted Moving Average

Weights **0.50, 0.30, 0.20** are applied to the three most recent months so that the latest observation receives the highest weight.

```{r wma}
wma_result <- method_results$weighted_moving_average
wma_result$weights
calc_accuracy_measures(wma_result$actual, wma_result$forecast)
```

**Interpretation:** Weighted averaging improves responsiveness to recent price changes compared with a simple moving average.

## 7.4 Exponential Smoothing

The smoothing parameter **alpha = 0.30** balances responsiveness and stability for a volatile monthly inflation-related index.

```{r es}
es_result <- method_results$exponential_smoothing
calc_accuracy_measures(es_result$actual, es_result$forecast)
```

**Interpretation:** Exponential smoothing gives more weight to recent observations in a recursive way and performs better than the naïve method when the latest level is informative.

## 7.5 Trend-Adjusted Exponential Smoothing

The series displays a meaningful upward trend, so Holt-style smoothing is applicable with **alpha = 0.30** and **beta = 0.20**.

```{r holt}
holt_result <- method_results$trend_adjusted_smoothing
calc_accuracy_measures(holt_result$actual, holt_result$forecast)
```

**Interpretation:** This method captures both level and trend, which is important for a producer price index under persistent inflation.

## 7.6 Linear Trend Projection

```{r linear-trend}
lt_result <- method_results$linear_trend
lt_result$equation
summary(lt_result$model)
calc_accuracy_measures(lt_result$actual, lt_result$forecast)
```

**Interpretation of intercept:** The intercept represents the estimated index level at the beginning of the sample period after accounting for the linear time trend.

**Interpretation of slope:** The positive slope indicates that the total producer price index increases over time on average.

## 7.7 Seasonal Indices

Monthly data support a 12-season structure.

```{r seasonal-indices}
si_result <- method_results$seasonal_indices
round(si_result$seasonal_index, 3)
calc_accuracy_measures(si_result$actual, si_result$forecast)
```

**Interpretation:** Months with seasonal indices above 1.00 tend to have relatively higher producer price index levels after removing trend effects.

## 7.8 Additive Decomposition

```{r additive}
add_result <- method_results$additive_decomposition
calc_accuracy_measures(add_result$actual, add_result$forecast)
```

**Interpretation:** Additive decomposition separates trend, seasonal, and random components. It is suitable because seasonal swings are roughly stable relative to the level of the series.

## 7.9 Multiplicative Decomposition

```{r multiplicative}
mult_result <- method_results$multiplicative_decomposition
calc_accuracy_measures(mult_result$actual, mult_result$forecast)
```

**Interpretation:** Multiplicative decomposition is also applicable because all index values are positive and seasonal effects can widen as the series level rises.

## 7.10 Regression with Trend and Seasonal Dummy Variables

```{r regression-seasonal}
reg_result <- method_results$regression_seasonal
summary(reg_result$model)
calc_accuracy_measures(reg_result$actual, reg_result$forecast)
```

**Interpretation:** The trend coefficient shows the average monthly change in the index, while seasonal dummy variables capture systematic month-specific deviations from trend.

# Forecast Accuracy Comparison

```{r comparison}
next_forecasts <- extract_next_forecasts(method_results, series_tbl)
comparison_tbl <- make_comparison_table(method_results, next_forecasts)
knitr::kable(comparison_tbl, digits = 4)
write.csv(comparison_tbl, "outputs/tables/accuracy_comparison.csv", row.names = FALSE)
```

# Actual vs Forecast Plots

```{r save-plots, include=FALSE}
save_all_method_plots(series_tbl, method_results)
```

The plots are saved under `outputs/figures/`. The notebook displays representative comparisons below.

```{r plot-grid, fig.cap="Actual vs forecast comparisons for selected methods.", out.width='100%'}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (res in list(naive_result, ma_result, holt_result, reg_result)) {
  plot_actual_vs_forecast(
    series_tbl = series_tbl,
    method_result = res,
    save_path = tempfile(),
    title = res$method
  )
}
par(mfrow = c(1, 1))
```

# Selection of the Superior Method

```{r superior-method}
superior <- select_superior_method(comparison_tbl, method_results)
superior$method
superior$reason
```

The superior method is selected using both numerical accuracy and suitability to the data structure. A method with the lowest MAD is preferred only if it also matches the observed monthly trend and seasonality of the producer price index.

# Final Next-Period Forecast

```{r final-forecast}
final_method_name <- superior$method
final_key <- superior$method_key
final_result <- method_results[[final_key]]
final_forecast_value <- next_forecasts[[final_key]]

final_tbl <- data.frame(
  selected_superior_method = final_method_name,
  date_of_data_access = as.character(access_date),
  latest_available_observation = target_info$latest_observation,
  forecast_target_period = target_info$forecast_target_period,
  forecasted_value = round(final_forecast_value, 4),
  mad = comparison_tbl$mad[comparison_tbl$method_key == final_key],
  mape = comparison_tbl$mape[comparison_tbl$method_key == final_key]
)

knitr::kable(final_tbl, digits = 4)
write.csv(final_tbl, "outputs/tables/final_forecast.csv", row.names = FALSE)

plot_superior_method(
  series_tbl = series_tbl,
  method_result = final_result,
  next_forecast = final_forecast_value,
  target_period = target_info$forecast_target_period
)
```

**Interpretation of the final forecast:** The selected model forecasts the Non-Domestic Producer Price Index for **`r target_info$forecast_target_period`** at **`r round(final_forecast_value, 2)`**. This suggests that imported producer price pressure is expected to continue along its recent trend, adjusted for monthly seasonality.

# Limitations

- The series is affected by exchange-rate pass-through, energy prices, and policy shocks that simple forecasting methods do not model explicitly.
- TÜİK may revise previously published index values.
- Structural breaks during high-inflation periods reduce the stability of historical patterns.
- The project uses only the index level series and does not include explanatory macroeconomic variables.

# Reproducibility Notes

1. Clone the public GitHub repository.
2. Open `forecasting_project.Rmd` in RStudio.
3. Install required packages if prompted in the setup chunk.
4. Knit the notebook to regenerate tables, figures, and the HTML report.
5. All TÜİK data are downloaded live through `tuikr`; no manually prepared data files are used.

# Author

- **Name:** Mehmet Zahid Kemik
- **Student Number:** 138722034
- **Course:** Quantitative Analysis
